---
title: "ETC5513-Assignment3"
format:
  pdf:
    toc: true
number-sections: true
editor: visual
---

```{r}
#| message: false
#| warning: false
#| echo: false
install.packages("Polychrome")
library(tidyverse)
library(ggplot2)
library(Polychrome)
library(scales)
library(kableExtra)
library(knitr)

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      echo = FALSE)
```

```{r}
# Import data and filtering the variables
epl_raw <- read.csv("data/epl_final.csv")

epl_data <- epl_raw %>%
  select(Season, HomeTeam, AwayTeam, FullTimeResult, HomeShots, AwayShots)


# Filter to only include most recent seasons
recent_seasons <- c("2015/16", "2016/17", "2017/18", "2018/19", "2019/20",
                    "2020/21", "2021/22", "2022/23", "2023/24", "2024/25")

epl_recent <- epl_data %>%
  filter(Season %in% recent_seasons)
```

## Executive Summary

This study examines whether a higher average number of shots per game leads to better team performance in the English Premier League (EPL). Using match-level data from the 2000/01 to 2024/25 seasons, we focus on the most recent decade to analyze the relationship between shot volume and win rate. Our results show a generally positive correlation—teams that shoot more often tend to win more matches. However, exceptions exist, suggesting that while offensive intensity helps, it does not guarantee success.

## Introduction

Over the past two decades, the English Premier League (EPL) has undergone significant tactical transformation. Many clubs have shifted from traditionally balanced or defensive systems toward more aggressive, attack-oriented approaches. This evolution is visible in match statistics, with an upward trend in the average number of shots taken per game. Shots are commonly used as a proxy for offensive intent, and their increase suggests a broader strategic emphasis on attacking play.

However, while modern football increasingly prioritizes creating scoring opportunities, it remains unclear whether more aggressive offensive styles consistently yield better outcomes. Do teams that take more shots necessarily win more matches, or can success also be achieved through more conservative or balanced approaches. This study seeks to explore the relationship between offensive output and team performance in the EPL. Specifically, we investigate whether a higher average number of shots per game correlates with a higher win rate across seasons. 


## Methodology

### Dataset Introduction

```{r}
#| label: fig-timeplot
#| fig-cap: "Timeplot of Average Shots per Season in the EPL (2000/01–2024/25)"

# Create variable for total number of shots each game
epl_processed_method <- epl_data %>%
  mutate(TotalShots = HomeShots + AwayShots)

# Summarise average shots per match by season
summary_by_season <- epl_processed_method %>%
  group_by(Season) %>%
  summarise(AvgShotsPerMatch = mean(TotalShots, na.rm = TRUE))

# Create line plot of total shots per game
ggplot(summary_by_season, aes(x = Season, y = AvgShotsPerMatch, group = 1)) +
  geom_line(color = "darkblue", size = 1.1) + 
  labs(
    x = "Season",
    y = "Average Shots per Match") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Using match-level data from the 2000/01 to 2024/25 EPL seasons, we examine the relationship between a team's average shots per game and its win rate. Focusing on the 2015/16–2024/25 seasons, we compute per-season team averages. As shown in Figure @fig-timeplot, shots per match have risen, motivating our analysis of whether offensive intensity improves the team's performance in the league.

We used match-level data sourced from [Kaggle](https://www.kaggle.com/datasets/marcohuiii/english-premier-league-epl-match-data-2000-2025/data), covering EPL matches from the 2000/01 season through to 2024/25. From the full dataset, we extracted six key variables related to match outcomes and shot frequency: `Season`, `HomeTeam`, `AwayTeam`, `FullTimeResult`, `HomeShots`, and `AwayShots`. After processing, the extracted dataset contained 9,380 observations across 25 seasons.

### Data Description and Analytical Approach

```{r}
#| label: tbl-variabletable
#| tbl-cap: Table of Variable Names for Constructed Dataset

# Create a description table
variable_description <- tibble::tibble(
  Variable = c("Season", "Team", "TotalMatches", "TotalShots", "TotalWins", 
               "AvgShotsPerGame", "WinRatePercent"),
  
  Description = c(
    "The English Premier League season (e.g., '2015/16')",
    "Name of the football team",
    "Total number of matches played by the team that season",
    "Total number of shots taken by the team across the season",
    "Total number of wins recorded by the team that season",
    "Average number of shots made by the team each game in that season",
    "Win rate percentage (TotalWins / TotalMatches * 100)"))

kable(variable_description)

```

For analysis, we focused on the 10 most recent seasons (2015/16 to 2024/25) to reflect current trends. Using this subset, we constructed a team-season-level dataset, where each observation represents one team’s performance in a given season. The final dataset includes 200 observations and 7 variables, described above in @tbl-variabletable.

We use **Win Rate Percent**, which is based only on number of wins, for our performance indicator because wins carry the most points in league standings and best indicate a team’s ability to outperform opponents.

To explore the relationship between shot volume and success, we created a scatter plot of **Win Rate Percent** against **Average Shots Per Game**. Each point represents a team’s season, allowing us to visually assess whether teams that shoot more frequently also tend to win more matches.

## Results

```{r}
# Get stats for home games
home_stats <- epl_recent %>%
  mutate(
    Team = HomeTeam,
    Shots = HomeShots,
    Win = ifelse(FullTimeResult == "H", 1, 0)
  ) %>%
  select(Season, Team, Shots, Win)

# Get stats for away games
away_stats <- epl_recent %>%
  mutate(
    Team = AwayTeam,
    Shots = AwayShots,
    Win = ifelse(FullTimeResult == "A", 1, 0)
  ) %>%
  select(Season, Team, Shots, Win)

# Combine datasets for home and away stats
team_match_data <- bind_rows(home_stats, away_stats)

team_season_summary <- team_match_data %>%
  group_by(Season, Team) %>%
  summarise(
    TotalMatches = n(),
    TotalShots = sum(Shots, na.rm = TRUE),
    TotalWins = sum(Win, na.rm = TRUE),
    AvgShotsPerGame = TotalShots / TotalMatches,
    WinRatePercent = 100 * TotalWins / TotalMatches
  ) %>%
  ungroup()
```

```{r}
#| label: fig-facetplot
#| fig-cap: "Average Shots vs Win Rate by Season"
# Facet Plot to compare the Average Shos taken vs Win rate per season by each team

# Number of Teams:
number_of_teams <- length(unique(team_season_summary$Team))
# Creating color palette:
distinct_team_colors <- createPalette(number_of_teams,  
                                      seedcolors = c("#000000", "#FFFFFF"))

# Assign team names as names of the color vector
names(distinct_team_colors) <- unique(team_season_summary$Team)

ggplot(
  team_season_summary,
  aes(x = AvgShotsPerGame, 
      y = WinRatePercent, 
      color = Team)) +
  geom_point(size = 2, alpha = 0.85) +
  facet_wrap(~ Season, scales = "free_x") +
  scale_color_manual(values = distinct_team_colors) +
   scale_x_continuous(labels = number_format(accuracy = 1)) +
  scale_y_continuous(labels = number_format(accuracy = 1)) +
  labs(
    title = "Average Shots vs Win Rate by Season",
    x = "Average Shots Per Game",
    y = "Win Rate (%)",
    color = "Team"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
    hjust = 1, size = 6
      ),
    axis.text.y =  element_text(
      hjust = 1, size = 6
      ),
    plot.title =  element_text(
      size = 10, face = "bold"
      ),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 6, face = "bold")
    )
```

The above visualization in @fig-facetplot, we can see the relationship between the average number of shots and the winning rate of the EPL Teams across multiple seasons. We have 34 unique teams in our data and we have given distinct colors for all the teams. The scatter plot is comparing the Average Shots Per Game vs Wining rate of the teams.

In @fig-facetplot, we can see that across all seasons, there is a positive trend : **teams that attempt more shots per game generally achieve higher win rates**. But, there are some case scenarios as well where a team has attempted more shots but have had a lower win rate, e.g., Liverpool in 2023/24 season attempted 21 shots on an average but had lesser win rate in comparison to Man City who have attempted only 18 shots on an average.

## Discussions, Conclusions & Recommendations

The findings from our visual analysis reveals that there is a positive relationship between the average shots and win rate. But, we can also see that there are several outliers as well which shows that there are some teams which even though attempting more shots did not achieve a higher win rate. This indicates that the number of shots a team makes is not the only factor that determines its performance in the league. The quantity of the shots may not be enough to give accurate results. We can look into the factors which specify the quality of shots as well.

This analysis demonstrate that EPL teams that produce more shots per game are more likely to achieve higher win rates. While the relationship is not absolute, the patter is strong enough to highlight that the offensive strategies for the teams are a good way for a season long success.

-   Further Analysis: We can include more variables like defensive errors, actual goals, quality of shots, etc. to improve our model.

-   Top Performers: We can also compare the outliers and other team (Top performing teams) to see the patterns of abnormality.

### References
